\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-11 k-planar.tex
Pffffft

Let $G$ be a $1$-plane graph.
% Then $G$ has $O(V(G)$ edges since the edges of $G$ can be partitioned into two planar graphs by placing one edge from each crossing pair into a second set.
% \footnote{We may assume that $G$ has no parallel edges, since removing all but one copy from each set of parallel edges does not change the set of adjacent pairs of vertices in $G$, so any supergraph that contains}
We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.
% This increases the number of edges of $G$, by $O(|V(G)|)$ since these newly added edges do not cross any other edges of $G$.

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that $V(F)$ can be partitioned into (possibly empty) sets $P_1,\ldots,P_3$ such that for each $i\in\{1,2,3\}$,
		\begin{compactenum}
			\item $F[P_i]$ is a path; and
			\item $|V(P_i)\cap L_j| \le 3$ for every integer $j\ge 0$.
		\end{compactenum}
% \note{DW}{Should the vertices already grabbed by $P_1,P_2,P_3$ be mentioned in the setup? The new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.} \note{PM}{No, at this point we're only promising an $H$-partition of $N$, the part of $G$ contained in the uncrossed cycle $F$.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x\cup\{v_x\} : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 3$;
		\item for each $i\in\{1,2,3\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$;
%		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}\note{PM}{No, for the same reason as above.}
 and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

\begin{proof}
	This proof is very similar to the proof of Lemma~13 by \citet{DJMMUW20}, but is sufficiently complicated by the $1$-planar setting to warrant a full proof.
   %
   %
   %
   %
   % Rather than duplicate every detail of that proof here, we focus on the differences and refer the reader to the original proof for the remaining details.
  In particular, the reader should keep in mind that $G$ and $G'$ are multigraphs and that the cycle $F$ may consist of two vertices and two (parallel) edges.

  The proof is by induction on the number of inner vertices of $N$, i.e., on $|V(N)\setminus V(F)|$.  If $N$ has no inner vertices, then we can simply take $\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements of the lemma.  Therefore, we may assume that $N$ has at least one inner vertex.

  If $k\in\{1,2\}$ then we let $\tau:=v_1,\ldots,v_p$ be an inner face of $N'$ that contains at least one edge of $F$ on its boundary.  If $k=3$ then assign each vertex $v$ of $N'$ a colour $\alpha(v)\in\{1,\ldots,3\}$ as follows:  Let $w$ be the first vertex of $F$ encountered on the path in $T$ from $v$ to the root of $T$.  (This implies that $w=v$ when $v$ is a vertex of $F$.)  Since $w\in V(F)$, $w\in P_i$ for some $i\in\{1,\ldots,k\}$ and we define $\alpha(v):=i$.  Now add edges to the inner faces of $N'$ to obtain a supergraph $N''$ in which each inner face has three $3$ edges on its boundary. By Sperner's Lemma \note{DW}{Does this work for multigraph near-triangulations?} there exists an inner face $\tau'':=v_1,v_2,v_3$ of $N''$ such that $\alpha(v_i)=i$ for each $i\in\{1,2,3\}$.  Let $\tau:=v_1,\ldots,v_p$ be the face of $N'$ that contains $\tau''$.  (Note that $p\in\{3,4\}$.)

  % If $k=2$, then let $\tau:=v_1v_2$ be any edge of $N'$ whose interior is in the interior of $F$ and whose endpoints have different colours.  The existence of such an edge is straightforward to establish by considering the auxilliary graph $A$ whose vertices are the faces of $N'$ and in which two faces are adjacent if and only if they share an edge whose endpoints have different colours.  Every vertex in $A$ has degree $0$ or $2$ and the outer face $F_0$ of $N'$ has degree $2$ in $A$.  Therefore $A$ contains a cycle $C$ that contains $F_0$.  Two edges of $C$ correspond to bichromatic edges of $F$ and the the remaining edges correspond to bichromatic edges in the interior of $F$, so it only remains to verify that $C$ has length at least $3$.  The only other possibility is that $C$ has length $2$, which

  % See, for example, \cite[Lemma~6]{bose.morin.ea:optimal}.

  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to $V(F)$.
   %
   %
   % If $k=3$, set $R_i := P_i$ for each $i\in\{1,2,3\}$.  Otherwise, as in \citep{DJMMUW20}, split $P_1,\ldots,P_k$ to partition $V(F)$ into three sets $R_1$, $R_2$, and $R_3$ such that each $F[R_i]$ is a non-empty path and each $R_i$ contains vertices from exactly one of $P_1,\ldots,P_k$.
  %
	% Next, as in \citep{DJMMUW20}, use Sperner's Lemma to find an inner face $\tau=v_1v_2v_3$ of $N'$ such that, $T$ contains disjoint vertical paths $Q_1,Q_2,Q_3$ such that each $Q_i$ begins at $v_i$, ends at some vertex in $R_i$, and whose internal vertices (if any) are contained in $N'-V(F)$.
  Let $\overline{Y}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $S:=V(Q_1)\cup V(Q_2)\cup V(Q_3)$ and let $v:=v_p$.  Observe that, for each $i\in\{1,2,3\}$ and each $j\ge 0$, $|V(Q_i)\cap L_j|\le 1$.  Therefore $|S\cap L_j|\le 3$ for each integer $j\ge 0$.  Therefore, taking $Y:=S\cup\{v\}\setminus V(F)$ as a part in the partition $\PP$ will satisfy Requirement~2 of the lemma.

>>>>

Let $G$ be a $1$-plane graph.  We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that $V(F)$ can be partitioned into (possibly empty) sets $P_1,\ldots,P_3$ such that for each $i\in\{1,2,3\}$,
		\begin{compactenum}
			\item $F[P_i]$ is a path; and
			\item $|V(P_i)\cap L_j| \le 3$ for every integer $j\ge 0$.
		\end{compactenum}
% \note{DW}{Should the vertices already grabbed by $P_1,P_2,P_3$ be mentioned in the setup? The new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.} \note{PM}{No, at this point we're only promising an $H$-partition of $N$, the part of $G$ contained in the uncrossed cycle $F$.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x\cup\{v_x\} : x\in V(H)\}$ such that:
	\begin{compactenum}[(i)]
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 3$;
		\item for each $i\in\{1,2,3\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$;
%		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}\note{PM}{No, for the same reason as above.}
 and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

\begin{proof}
  This proof is very similar to the proof of Lemma~13 by \citet{DJMMUW20}, but is sufficiently complicated by the $1$-planar setting to warrant a full proof.  In particular, the reader should keep in mind that $G$ and $G'$ are multigraphs and that the cycle $F$ may consist of two vertices and two (parallel) edges.

  The proof is by induction on the number of inner faces of $N'$.  If $N'$ has only one inner face then we can simply take $\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements (i)--(iv):
  \begin{inparaenum}[(i)]
    \item $H:=N/\mathcal{P}$ is planar because any graph on $k\le 3$ vertices is planar.
    \item Assumption 4(a) ensures that $|S_x\cap L_i|\le 3$ for each $x\in V(H)$.
    \item By definition, each each of $P_1,\ldots,P_k$ is s part of $\mathcal{P}$.
    \item The trivial tree decomposition of $H$ that places $x_1,\ldots,x_k$ in one bag satisfies this requirement.
  \end{inparaenum}
  Therefore, we may assume that $N$ has more than one inner face.

  The $H$-partition $\mathcal{P}$ will include $P_1$, $P_2$ , $P_3$ and a part $Y$, called a \emph{tripod}, that we now define.  The tripod $Y$ is determined by three vertices $v_1v_2v_3$ that belong to a common inner face $\tau$ of $N'$.  Depending on the value $k$, there are two cases to consider:
  \begin{compactenum}
    \item If $k\in\{1,2\}$ then we let $\tau:=v_1,\ldots,v_p$ be an inner face of $N'$ that contains at least one edge of $F$ on its boundary.

    \item If $k=3$ then assign each vertex $v$ of $N'$ a colour $\alpha(v)\in\{1,\ldots,3\}$ as follows:  Let $w$ be the first vertex of $F$ encountered on the path in $T$ from $v$ to the root of $T$.  (This implies that $w=v$ when $v$ is a vertex of $F$.)  Since $w\in V(F)$, $w\in P_i$ for some $i\in\{1,\ldots,k\}$ and we define $\alpha(v):=i$.  Now add edges to the inner faces of $N'$ to obtain a supergraph $N''$ in which each inner face has three $3$ edges on its boundary. By Sperner's Lemma \todo{DW:Does this work for multigraph near-triangulations?} there exists an inner face $\tau'':=v_1,v_2,v_3$ of $N''$ such that $\alpha(v_i)=i$ for each $i\in\{1,2,3\}$.  Let $\tau:=v_1,\ldots,v_p$ be the face of $N'$ that contains $\tau''$.  (Note that $p\in\{3,4\}$.)
  \end{compactenum}

  \todo[inline]{
    Answering David's question: Yes, it only requires that the inner faces have three sides and that the outer face is bounded by a simple cycle.  The proof goes through without change.  Look at the subgraph $X$ of the dual that contains only the edges dual to red/blue edges.  In our setting, the outer face has degree $1$ in $X$, which is odd.  Any bichromatic inner face has degree $0$ or degree $2$ in $X$.  There has to be another vertex of odd degree somewhere, which can only come from a trichromatic inner face.
  }

  We can now define the vertices in the tripod $Y$.  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to $V(F)$.  Let $\overline{Y}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $S:=V(Q_1)\cup V(Q_2)\cup V(Q_3)$ and let $v:=v_p$.  Observe that, for each $i\in\{1,2,3\}$ and each $j\ge 0$, $|V(Q_i)\cap L_j|\le 1$.  Therefore $|S\cap L_j|\le 3$ for each integer $j\ge 0$.  Therefore, taking $Y:=S\cup\{v\}\setminus V(F)$ as a part in the partition $\PP$ will not violate Condition~(ii).
